Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany

### 2025-06-10 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
paper done

---
title: "Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany"
title-running: Validation challenges in tree crown segmentations
format:
  quarto-ecml-pdf:
    keep-tex: true
  quarto-ecml-html: default
author:
  - name: Taimur Khan
    corresponding: true 
    affiliations:
      - name: Helmholtz Centre for Environmental Research (UFZ)
        address: Theodor-Lieser-Str. 4, 06120 Halle (Saale)
        country: Germany
        url: \{taimur.khan\}\@ufz.de
        id: south
    url: https://www.ufz.de/index.php?en=49404
  - name: Calvin Cordozar Broadus Jr.
    affiliations:
      - name: Fictional West Coast University
        address: Long Beach CA 90840
        country: USA
        url: ccb@fwcu.fake
      - name: Secondary European Affiliation
        address: Tiergartenstr. 17
        city: Heidelberg
        country: Germany
        postal-code: 69121 
        url: lncs@springer.com
  - name: Antwan André Patton
    orcid: '0000-1111-2222-3333'
    affiliations:
      - ref: south 
author-running: T. Khan et al.
abstract: |
 Recent progress in deep learning has greatly improved the ability to extract detailed information from remote sensing images, such as identifying and outlining individual tree crowns. However, as these models are scaled up for use across large geographic areas—beyond individual forests or research plots—new challenges begin to emerge. One of the main hurdles is validating how accurately these models detect and segment tree crowns. Even small errors in these predictions can affect decisions in forestry management and urban planning, limiting the real-world usefulness of such models. Our experience working with large-scale applications across regions in Germany highlights how factors like tree size, seasonal variations, shadow effects, and the shape of training data annotations influence model performance. This underlines the importance of thoughtful selection and design of training datasets. We discuss these challenges in validation and emphasize the need for careful model evaluation when working at broader spatial scales.
keywords: [First Keyword, Second Keyword, Another Keyword.]
---

@@

The integration of deep learning into remote sensing has opened new ways in how we monitor and analyse ecological systems [@zheng2024review; @zhao2023systematic]. Tree crown segmentation has especially benefited from advances in convolutional neural networks (CNNs) and transformer-based architectures. Latter was used by [@JIANG2025103114], for single tree detection. Approaches leveraging high-resolution imagery have demonstrated remarkable capabilities in delineating individual tree crowns [@weinstein2019individual; @khant2025deeptrees; @freudenberg2022individual], illustrating the potential for scalable and automated tree mapping.

However, while these deep learning models demonstrate impressive performance in controlled environments or with limited-scale datasets, their deployment at expansive spatial extents, such as regional or national scales, exposes a distinct spectrum of challenges that remain largely unaddressed. Foremost among these is validation—a critical yet underexplored hurdle in ensuring the reliability and generalizability of model outputs. As recent studies have shown, the performance of tree crown segmentation models is highly sensitive to the characteristics of the training data, such as tree size, species diversity, seasonal variations, vitality conditions, and image quality [@moussaid2021tree; @cong2022citrus], a fact that directly reflects the profound complexity of how tree crowns appear in real-world conditions. To enable scalable validation frameworks, it is therefore necessary to accurately characterize and account for this complexity.

@@

![]("./figures/figure1.png")

@@

![]("./figures/figure2.png")

*Figure 2:* Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue & Sachsen; in green) in Germany. Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps. Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.

@@

# Conclusion
@
Deep learning has opened a new frontier for tree crown segmentation from remote sensing imagery, yet its promise is contingent on our ability to validate these models reliably and at scale. As this paper has outlined, the challenges are not merely technical—they are epistemological. They force us to ask: what does it mean to “know” a tree from above, when the canopy is a moving target shaped by phenology, terrain, land cover and image artifacts?

@@

![]("./figures/figure3.png") 
*Figure 3:*

*Table 1: Summary of key validation challenges in large-scale tree crown segmentation, with examples, relevant references, and recommended mitigation strategies.*

@@

>>>>

---
title: "Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany"
title-running: Validation challenges in tree crown segmentations
bibliography: bibliography.bib
format:
  quarto-ecml-pdf:
    keep-tex: true
  quarto-ecml-html: default
author:
  - name: Taimur Khan
    orcid: '0000-0001-7833-5474'
    corresponding: true 
    affiliations:
      - name: Helmholtz Centre for Environmental Research (UFZ)
        address: Theodor-Lieser-Str. 4
        city: Halle (Saale)
        postal-code: 04109 
        country: Germany
        url: taimur.khan@ufz.de
        id: south
    url: https://www.ufz.de/index.php?en=49404
  - name: Jasmin Krebs
    orcid: '0009-0003-9330-1231'
    affiliations:
      - name: Leipzig University
        address: Augustusplatz 10 
        city: Leipzig
        country: Germany
        postal-code: 04109 
        url: jassi.krebs@gmx.de
  - name: Sharad Kumar Gupta
    orcid: '0000-1111-2222-3333'
    affiliations:
      - ref: south
      - name: Secondary European Affiliation
        address: Tiergartenstr. 17
        city: Heidelberg
        country: Germany
        postal-code: 69121 
        url: sharad.gupta@ufz.de
  - name: Jonathan Renkel
    orcid: '0000-1111-2222-3333'
    affiliations:
      - name: Martin Luther University Halle-Wittenberg
        address: Von-Seckendorff-Platz 2
        city: Halle (Saale)
        country: Germany
        postal-code: 06120
        url: jonathan.renkel@geo.uni-halle.de
  - name: Caroline Arnold
    orcid: '0000-0002-9458-1517'
    affiliations:
      - name: Helmholtz-Zentrum Hereon
        address: Max-Planck-Straße 1
        city: Geesthacht
        country: Germany
        postal-code: 21502 
        url: caroline.arnold@hereon.de
  - name: Nils Nölke 
    orcid: '0000-0003-4925-2287'
    affiliations:
      - name: University of Göttingen, Forest Inventory and Remote Sensing
        address: Büsgenweg 5
        city: Göttingen
        country: Germany
        postal-code: 37077 
        url: nils.noelke@forst.uni-goettingen.de

author-running: T. Khan et al.
abstract: |
  Recent advancements in deep learning have significantly enhanced the extraction of detailed ecological information from multispectral (RGBi) aerial imagery, notably the identification and segmentation of individual tree crowns. Using data from our extensive case study mapping approximately 218.7 million trees across the German federal states of Sachsen and Sachsen-Anhalt, we illustrate how scaling these models beyond controlled settings introduces critical validation challenges. Even minor inaccuracies in tree crown segmentation can substantially impact practical applications in forestry management and urban planning. Our findings emphasize the influence of tree allometry, seasonal variability, shadow effects, and the characteristics of training dataset annotations on model accuracy. Addressing these issues necessitates careful selection and design of training data and exploration of task-specific pre-training through Foundation Models. We highlight these validation complexities and recommend for rigorous steps to ensure the reliability and practical utility of large-scale deep learning models in ecological and urban management contexts.
keywords: [Deep Learning, Ecology, Forestry, Remote Sensing.]
---

@@

The integration of deep learning into remote sensing has opened new ways in how we monitor and analyse ecological systems [@zheng2024review; @zhao2023systematic]. Tree crown segmentation has especially benefited from advances in convolutional neural networks (CNNs) and transformer-based architectures. Latter was used by [@JIANG2025103114], for single tree detection. Approaches leveraging high-resolution imagery have demonstrated remarkable capabilities in delineating individual tree crowns [@weinstein2019individual; @khan2025deeptrees; @freudenberg2022individual, @JIANG2025103114], illustrating the potential for scalable and automated tree mapping.

However, while these deep learning models demonstrate impressive performance in controlled environments or with limited-scale datasets, their deployment at expansive spatial extents, such as regional or national scales, exposes a distinct spectrum of challenges that remain largely unaddressed. Foremost among these is validation — a critical yet underexplored hurdle in ensuring the reliability and generalizability of model outputs. As recent studies have shown, the performance of tree crown segmentation models is highly sensitive to the characteristics of the training data, such as tree size, species diversity, seasonal variations, vitality conditions, and image quality [@moussaid2021tree; @cong2022citrus], a fact that directly reflects the profound complexity of how tree crowns appear in real-world conditions. To enable scalable validation frameworks, it is therefore necessary to accurately characterize and account for this complexity.

@@

![a. Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns.
b. Relationship between normalized crown area (x-axis) and estimated crown spread (y-axis) for all segmented trees. A clear positive correlation is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns. This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.]("./figures/figure1.png")

@@

![Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue & Sachsen; in green) in Germany. Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps. Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.]("./figures/figure2.png")

@@

# Conclusion

Deep learning has opened a new frontier for tree crown segmentation from remote sensing imagery, yet its promise is contingent on our ability to validate these models reliably and at scale. As this paper has outlined, the challenges are not merely technical—they are epistemological. They force us to ask: what does it mean to “know” a tree from above, when the canopy is a moving target shaped by phenology, terrain, land cover and image artifacts?

@@

![Environmental heterogeneity across the German federal states of Sachsen and Sachsen-Anhalt, where tree crown segmentation was conducted.
(Top) Acquisition months of DOP20 imagery (April to September) used in the segmentation, showing strong spatial variation in image phenology ( [@geosn_dop20_2022]; [@lvermgeo2020dop]).
(Middle) Aggregated land cover classes based on CORINE Land Cover 2020 data, grouped into five categories: agricultural areas, artificial surfaces, forest and semi-natural areas, water bodies, and wetlands ( [@eea_corine2020]).
(Bottom) Elevation map derived from the COPDEM GLO-30 digital elevation model at 30 m spatial resolution, representing terrain variability used to assess segmentation accuracy across topographic gradients ( [@copernicus_copdem30]).]("./figures/figure3hor.png")

@@

: Summary of key validation challenges in large-scale tree crown segmentation, with examples, relevant references, and recommended mitigation strategies. {#tbl-headings .tbl-colwidths="15,25,20,10,20,10"}
